# Introduction to the spsann package

## Objective functions

### Points (or point-pairs) per lag distance class - PPL

#### Distances

Euclidean distances between points are calculated. This computation requires
the coordinates to be projected. The user is responsible for making sure that
this requirement is attained.

#### Distribution

Using the default uniform distribution means that the number of **point-pairs**
per lag distance class is equal to $n \times (n - 1) / (2 \times lag)$, where
$n$ is the total number of points in `points`, and $lag$ is the number of lag
distance classes.

Using the default uniform distribution means that the number of **points** per
lag distance class is equal to the total number of points in `points`. This is
the same as expecting that each point contributes to every lag distance class.

Distributions other that the available options can be easily implemented 
changing the arguments `lags`, `lags.base` and `pre.distri`.

#### Type of lags

Two types of lag distance classes can be created by default. The first 
(`lags.type = "equidistant"`), are evenly spaced lags. They are created by
simply dividing the distance interval from 0.0001 to `cutoff` by the required
number of lags. The minimum value of 0.0001 guarantees that a point does not
form a pair with itself.

The second type (`lags.type = "exponential"`) of lag distance classes is defined
by exponential spacings. The spacings are defined by the base $b$ of the
exponential expression $b^n$, where $n$ is the required number of lags. The base
is defined using argument `lags.base`. For example, the default `lags.base = 2`
creates lags that are sequentially defined as half of the immediately preceding
larger lag. If `cutoff = 100` and `lags = 4`, the upper limits of the lag
distance classes will be

```
> 100 / (2 ^ c(1:4))
[1] 50.00 25.00 12.50  6.25
```
#### Criteria

The functions `objPoints` and `objPairs` (to be implemented) were designed to be
used in spatial simulated annealing to optimize spatial sample configurations.
Both of them have two criteria implemented. The first is called using 
`criterion = "distribution"` and is used to minimize the sum of differences
between a pre-specified distribution and the observed distribution of points or
point-pairs per lag distance class.

Consider that we aim at having the following distribution of points per lag: 

```{r}
desired <- c(10, 10, 10, 10, 10)
```

and that the observed distribution of points per lag is the following:

```{r}
observed <- c(1, 2, 5, 10, 10)
```

The objective at each iteration of the optimization will be to match the two 
distributions. This criterion is of the same type as the one proposed by 
Warrick and Myers (1987).

The second criterion is called using `criterion = "minimum"`. It corresponds to
maximizing the minimum number of points or point-pairs observed over all lag
distance classes. Consider we observe the following distribution of points per
lag in the first iteration:

```{r}
observed <- c(1, 2, 5, 10, 10)
```

The objective in the next iteration will be to increase the number of points
in the first lag ($n = 1$). Consider we then have the following resulting
distribution:

```{r}
resulting <- c(5, 2, 5, 10, 10)
```

Now the objective will be to increse the number of points in the second lag
($n = 2$). The optimization continues until it is not possible to increase the
number of points in any of the lags, that is, when:

```{r}
distribution <- c(10, 10, 10, 10, 10)
```

This shows that the result of using `criterion = "minimum"` is similar to using
`criterion = "distribution"`. However, the resulting sample pattern can be
significantly different. The running time of each iteration can be a bit longer
when using `criterion = "distribution"`, but since it is a more sensitive
criteriom (it takes all lags into account), convergence is likely to be
attained with a smaller number of iterations. Note that this also depends on
the other parameters passed to the optimization algorithm.

It is important to note that using the first criterion (`"distribution"`) in
simulated annealing corresponds to a **minimization** problem. On the other 
hand, using the second criterion (`"minimum"`) would correspond to a 
**maximization** problem. We solve this inconsistency substituting the criterion
that has to be maximized by its inverse. For conveninence we multiply the
resulting value by a constant (i.e. $c / x + 1$, where $c$ is the number of
points and $x$ is the criterion value). This procedure allows us to define both
problems as minimization problems.

#### Utopia and nadir points

Knowledge of the utopia and nadir points can help in the construction of 
multi-objective optimization problems.

When `criterion = "distribution"`, the **utopia** ($f^{\circ}_{i}$) point is
exactly zero ($f^{\circ}_{i} = 0$). When `criterion = "minimum"`, the utopia
point is approximately 1 (0.9) ($f^{\circ}_{i} \sim 1$). It can be calculated
using the equation $n / n + 1$, where $n$ is the number of points (`objPoints`),
or the number point-pairs divided by the number of lag distance classes
(`objPairs`).

The **nadir** ($f^{max}_{i}$) point depends on a series of elements. For
instance, when `criterion = "distribution"`, if the desired distribution of
point or point-pairs per lag distance class is 

```{r}
pre.distribution <- c(10, 10, 10, 10, 10)
```

the worst case scenario would be to have all points or point-pairs in a single
lag distance class, that is, 

```{r}
obs.distribution <- c(0, 50, 0, 0, 0)
```

In this case, the nadir point is equal to the sum of the differences between the
two distributions:

```
sum((c(10, 10, 10, 10, 10) - c(0, 50, 0, 0, 0)) ^ 2) = 2000
```

When `objective = "minimum"`, the nadir point is equal to 
$f^{max}_{i} = n / 0 + 1 = n$.

### Association/correlation and marginal distribution of the covariates - ACDC

### Mean squared shortest distance - MSSD

### Mean (or maximum) universal kriging variance - MUKV
